\chapter*{Preface\label{cha:preface}}

\epigraphfontsize{\small\itshape}
\epigraph{''Habiendo accedido tarde a los estudios, sentía ahora urgencia por estudiar. A veces, inmerso en sus libros, le venía a la cabeza la conciencia de todo lo que no había leído y la serenidad con la que trabajaba se hacía trizas cuando caía en la cuenta del poco tiempo que tenía en la vida para aprender tantas cosas, para aprender todo lo que tenía que saber''}{--- \textup{Stoner}}

Applied science has made societies evolved thanks to the existing linked between the basic science and the inventions, which implement the structural and basic knowdlege into the daily reality of the individuals. The bridge between the fundamental studies and its evolution or transformation into something useful for the societies development, has been a matter of interpretation or ``translation'' to find the accurate language to exchange this knowdlege.

% Esta transversalidad entre las disciplinas básicas de la física, la química o la biología con distintas acepciones prácticas, la mayoría de ellas recogidas bajo el paraguas de las ingenierías, desemboca ineludiblemente en una transformación de lo abstracto en lo tangible con la premisa o el ideal de mejorar el bienestar de las sociedades presentes y futuras.

Transversality between the fundamental disciplines of physiscs, chemistry or biology, and the different applied sciences, most of them under the umbrella of engineering, end unavoidably into a transformation of the abstraction into the tanglible with the premise or ideal of improving wellbeing of present and future societies.

% Es sin embargo muy probable, que esta concepción de la aplicación científica haya desembocado en el mayor problema al que la humanidad tiene que hacer frente desde su existencia, el cambio climático. Siendo este antropocentrismo parte del problema, y no de la solución. Jorge Wasenberg escibe en su libro, ``el pensador intruso'' en referencia a la evolución de la ciencia y el progreso subyacente lo siguiente:

However, it is very likely that this conception of the scientific application and what is called ``developement'' had led at the same time to the most challeging problems that humanity has to face from its existance, climate change, pollution, evironmental degradation, loss of biodiversity etc. The anthropocentrism have become a problem and all the science and technology that made better human lifes should be re-thought.


The inter-disciplinary and multi-disciplinary approaches are the only way to solve the complex problems that we will have to face. In his book, ``El pensador intruso'', where boundaries between knowdlege disciplines are explored, Jorge Wasengber wrote about the evolution of science and the underlying progress as follows:\\ 

%'There are above all two vices that tend to set 'pre-cooked' ideollogy into science. First is base on different kinds of anthropocentrism and consist in put the knowdlege subjet into the cosmos's center. The history of knowdlege is the witness: each time we 'barremos' the 'I' from the spot, the knowdlege progresses and only because of it.'
\textit{Existen sobre todo dos vicios que tienden a inyectar ideología precocinada en la ciencia. Una de ellas se basa en las distintas formas de antropocentrismo y consiste en situar instintivamente al sujeto del conocimiento en el centro del cosmos. La historia del conocimiento es testigo: cada vez que barremos el Yo del centro del escenario el conocimiento avanza, y avanza sólo por ello.}\\

Despite of the character of most applied science about solving ``just'' humanity problems, an holistic approach to these challenges would undoubtely lead to a better, more sustainable and responsible solutions, clearly showing the limitations and consequencies of our propositions. In a complex world, with complex societies and problems, answers must come from deep scientific reflection.

% We now need to allow Science to evolve and find the solution to our problems 
% En la coyuntura actual, el ser humano debe alejarse de la solucion antropogénica. Debemos ser parte de la solucion como ector de cambio, sin caer en el egoísmo que hasta hora nos ha conducido hasta aquí.
% Sin caer en lo pretencioso, este trabajo contribuye a la imbricación entre la ciencia del cambio climático y la tecnología renovable de producción eléctrica con mayor potencial a día de hoy, la solar fotovoltaica.

